The dynamics of starvation and recovery

### 2017-12-13 NatureComm_REVISION/YKR_NatureComm-Revision-4.tex
editorial revisions

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. 
Of particular importance is the interplay between reproduction and the dynamics of starvation and recovery. % in resource-limited environments
We introduce a nutritional state-structured model that incorporates two classes of consumer: nutritionally replete, reproducing consumers, and undernourished, non-reproducing consumers. % that are susceptible to mortality
% As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. To elucidate the consequences of this tradeoff, w
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics are typically driven to a steady state. %subject to these constraints between body size and a variety of traits
Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized. 
% Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. 
We also show that our model provides a natural framework that predicts
maximum body size for mammals by determining the relative stability of an
otherwise homogeneous population to a competing population with altered percent
body fat, providing a principled mechanism for a within-lineage driver of
Cope's rule.
% For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}

@@

The behavioral ecology of all organisms is influenced by the energetic state of individuals, which directly impacts how they invest reserves in uncertain environments.
Such behaviors are generally manifested as tradeoffs between investing in somatic maintenance and growth, or allocating energy towards reproduction~\citep{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.
The timing of these behaviors responds to selective pressure, as the choice of the investment impacts future fitness~\citep{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.
The influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to repeated delays or shifts in reproduction over the course of an organism's life.

The balance between (a) somatic growth and maintenance, and (b) reproduction depends on resource availability~\citep{Morris:1987eo}.
For example, reindeer invest less in calves born after harsh winters (when the mother's energetic state is depleted) than in calves born after moderate winters~\citep{Tveraa:2003fq}.
Many bird species invest differently in broods during periods of resource scarcity compared to normal periods~\citep{Daan:1988va,Jacot:2009dv}, sometimes delaying or even foregoing reproduction for a breeding season~\citep{Martin:1987dl,Stearns:1989ip,Barboza:2002in}.
Even freshwater and marine zooplankton have been observed to avoid reproduction under nutritional stress~\citep{Threlkeld:1976ih}, and those that do reproduce have lower survival rates~\citep{Kirk:1997cc}.
Organisms may also separate maintenance and growth from reproduction over space and time: many salmonids, birds, and some mammals return to migratory breeding grounds to reproduce after one or multiple seasons in resource-rich environments where they accumulate reserves~\citep{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.

@@

Including individual energetic dynamics~\citep{Kooi2000} in a population-level framework~\citep{Kooi2000,Sousa:2010ez} is challenging~\citep{Diekmann:2010da}, and
a common simplifying approach is provided in the classic Lotka-Volterra (LV) model, which assumes that consumer population growth rate depends linearly on resource
  density~\citep{murdoch:2003}. Here, we introduce an alternative
  approach---the Nutritional State-structured Model (NSM)---that accounts for
  resource limitation via explicit starvation. In contrast to
  the LV model, the NSM incorporates two consumer states:
  hungry and full, with only the former susceptible to mortality and only the
  latter possessing sufficient energetic reserves to reproduce.
  Additionally, we incorporate allometrically derived constraints on reproduction \citep{Kempes:2012hy}, incorporating the timescales of starvation and recovery. 
  As we shall show, our model makes several important predictions: 
  (i) the dynamics are typically driven to a refuge far from cyclic behavior and extinction risk, 
  (ii) steady state conditions of the NSM accurately predict the measured biomass densities for mammals described by Damuth's law,
  %  energetic equivalence principle, 
  (iii) there is an allometrically constrained upper-bound for mammalian body size, and
  (iv) the NSM provides a selective mechanism for the evolution of larger body size, known as Cope's rule.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{fig_FixedPoint2.eps}
\caption{\small{ The transcritical (TC; dashed line) and Hopf bifurcation (solid line) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
  bifurcation conditions separate parameter space into unphysical (left of the TC), cyclic,
  and steady state dynamic regimes.  The colors show the steady state densities for the energetically replete consumers $F^*$.  
  % Steady state densities for the energetically deficient consumers $H^*$ (not shown)
  % scale with those for $F^*$.  
}\label{fig:fp}}
\end{figure}

Consumers transition from the full state $F$ to the hungry state $H$ at a
rate $\sigma$---the starvation rate---and also in proportion to the absence
of resources $(1-R)$ (we assume a maximum density, the carrying capacity equal to 1).  Conversely, consumers recover from state $H$ to state
$F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$ represents a
ratio between maximal resource consumption and the carrying capacity of the
resource. %and $\approx 2$.
The resources that are eaten by hungry consumers (at rate $\rho R + \delta$)
account for their somatic growth ($\rho R$) and maintenance ($\delta$).  Full
consumers eat resources at a constant rate $\beta$ that accounts for maximal
maintenance and somatic growth (see the SI for mechanistic
derivations of these rates from resource energetics).
%This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
The NSM represents an ecologically motivated fundamental extension of the idealized starving random walk model of foraging, which focuses on resource depletion, to include reproduction and resource replenishment~\citep{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is a more general formulation than previous models incorporating starvation~\citep{Persson:1998hz}.

@@

This system of nondimensional equations follows from a set of first-principle
relationships for resource consumption and growth (see the SI for a full derivation and the dimensional form).
Notice that the total consumer density $F+H$ evolves according to $\dot{F}+\dot{H}=\lambda F-\mu H$.
This resembles the equation of motion for the predator density in the LV model~\citep{murray2011mathematical}, except that the resource density does not appear in the growth term.
% As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.
The rate of reproduction is independent of resource density because it is
assumed that the satiated state of the full consumer allows it to partition a
constant amount of energy towards reproduction, whereas a starved consumer
partitions no energy towards reproduction.  The rate of reproduction for the
total consumer density \emph{is} dependent on resource density, which
determines the size of the full and starved portions of the consumer
population.  Similarly, the consumer maintenance terms ($\delta H$ and
$\beta F$) are independent of resource density because they represent a
minimal energetic requirement for consumers in the $H$ and $F$ state,
respectively.  It follows that model predictions are robust only when $R$ is
of the order of 1, which holds for all cases that we explore.

\noindent \paragraph*{ {\bf Steady states of the NSM.}} From the solution to the single internal fixed point (Eq.~\eqref{eq:ss}, see Methods), an obvious constraint on the NSM is that the reproduction rate $\lambda$ must be less than the starvation rate $\sigma$, so that the consumer and resource densities are positive.
%In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ thus represents a transcritical (TC) bifurcation~\citep{Strogatz:2008wo} that demarcates a physical from an unphysical regime where all steady-state densities become negative after intersecting the trivial fixed point $(F^*,H^*,R^*)=(0,0,0)$.
The biological implication of the constraint $\lambda<\sigma$ has a simple interpretation---the rate at which a macroscopic organism loses mass due to lack of resources is generally much faster than the rate of reproduction.
As we will discuss below, this inequality is a natural consequence of allometric constraints~\citep{Kempes:2012hy} for organisms within empirically observed body size ranges. %(Fig.~\ref{fig:gvs})

@@

When the starvation rate $\sigma\gg\lambda$, a substantial fraction of the consumers are driven to the hungry non-reproducing state.
Because reproduction is inhibited, there is a low steady-state consumer density and a high steady-state resource density.
However, if $\sigma/\lambda\to 1$ from above, the population is overloaded with energetically-replete (reproducing) individuals, thereby promoting transient oscillations between the consumer and resource densities (Fig.~\ref{fig:fp}).
If the starvation rate is low enough that the Hopf bifurcation is crossed, these oscillations become stable over time.
This threshold occurs at higher values of the starvation rate as the recovery rate $\rho$ increases, such that the range of parameter space giving rise to cyclic dynamics also increases with higher recovery rates.\\

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{Growth-trajectory-diagram.eps}
\caption{\small{ The growth trajectory over absolute time of an individual organism as a function of body mass.  
Initial growth follows the black trajectory to an energetically replete reproductive adult mass $m=\epsilon_\lambda M$. %which we assume is 95\% asymptotic mass $M$.  
Starvation follows the red trajectory to $m = \epsilon_\sigma \epsilon_\lambda  M$. 
Recovery follows the green curve to the replete adult mass, where this trajectory differs from the original growth because only fat is being regrown which requires a longer time to reach $\epsilon_\lambda M$. %different energetics and 
Alternatively, death from starvation follows the blue trajectory to $m=\epsilon_\mu \epsilon_\lambda  M$.}\label{fig:growth}}
\end{figure}

% \noindent {\bf Role of allometry} \\
%[Link Allometry stuff to our model - what does it provide to the story]
%[Introduce biological and linked constraints]
%[Allometries capture vast amounts of diversity via a single parameter --- body size]
%[Allometries have captured intterest etc across multiple scales and biological classes of organisms]
\noindent {\bf Results}
\noindent \paragraph*{{\bf The allometry of extinction risk.}} While there are no {\it a priori} constraints on the parameters in the NSM, we expect that each species should be restricted to a distinct portion of the parameter space.  We
use allometric scaling relations to constrain the covariation of rates in a
principled and biologically meaningful manner (see Methods).  Allometric scaling relations
highlight common constraints and average trends across large ranges in body
size and species diversity. Many of these relations can be derived from a
small set of assumptions and in the Methods we describe our framework to determine the
covariation of timescales and rates across a range of body sizes for each of
the key parameters of our model (cf. ref.~\citep{Yodzis:1992hg}).  
% We are thereby able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

Nearly all of the rates described in the NSM are determined by consumer
metabolism, which can be used to describe a variety of organismal features
\citep{Brown:2004wq}.
We derive relationships for the rates of reproduction, starvation, recovery, and mortality based on first principles, and as a function of an organism's body size and metabolic rate (see Methods).
Because we aim to explore the starvation-recovery dynamics as a function of an organism's body mass $M$, we parameterize these rates in terms of the \emph{percent} gain and loss of the asymptotic (maximum) body mass, $\epsilon M$, where different values of $\epsilon$ define different states of the consumer (Fig.~\ref{fig:growth}; see Methods for derivations of allometrically constrained rate equations).
Although the rate equations \eqref{eq:system} are general and can in
principle be used to explore the starvation recovery dynamics for most
organisms, here we focus on allometric relationships for terrestrial-bound
lower trophic level endotherms (see the SI for values), specifically herbivorous mammals, which range from a minimum of $M\approx1$g (the Etruscan shrew \emph{Suncus etruscus}) to a maximum of $M\approx10^7$g (the early Oligocene Indricotheriinae and the Miocene Deinotheriinae).
Investigating other classes of organisms would simply involve altering the metabolic exponents and scalings associate with $\epsilon$. Moreover, we emphasize that our allometric equations describe mean relationships, and do not account for the (sometimes considerable) variance associated with individual species.
{\color{red} We note that including additional allometrically-scaled mortality terms to both $F$ and $H$ does not change the form of our model nor impact our quantitative findings (see SI for derivation).}

@@

%Extinction to the left and the right
% \noindent \paragraph*{{\bf Extinction risk}} 
As the allometric derivations of the NSM rate laws reveal (see Methods), starvation and recovery rates are not independent parameters, and the biologically relevant portion of the phase space shown in Fig.~\ref{fig:fp} is constrained via covarying parameters.
Given the parameters of terrestrial endotherms, we find that the starvation rate $\sigma$ and the recovery rate $\rho$ are constrained to lie within a small region of potential values for the known range of body sizes $M$.
Indeed, starvation and recovery rates across all values of $M$ fall squarely in the steady state region at some distance from the Hopf bifurcation. 
This suggests that cyclic population dynamics should be rare, particularly in environments where resources are limiting.
% Moreover, the dynamics for all mammalian body sizes are confined to the steady-state regime of the NSM and that limit-cycle behavior is precluded.

Higher rates of starvation result in a larger flux of the population to the hungry state.
In this state, reproduction is absent, thus increasing the likelihood of extinction.  From the perspective of population survival, it is the rate of starvation relative to the rate of recovery that determines the long-term dynamics of the various species (Fig.~\ref{fig:fp}).
We therefore examine the competing effects of cyclic dynamics vs. changes in steady-state density on extinction risk, both as functions of $\sigma$ and $\rho$.
To this end, we computed the probability of extinction, where we define extinction as a population trajectory falling below one fifth of the allometrically constrained steady state at any time between $t=10^8$ and $t=10^{10}$.
This procedure was repeated for 50 replicates of the continuous-time system shown in Eq.~\ref{eq:system} for organisms with mass ranging from $10^2$ to $10^6$ grams.
%the initial condition isdistributed around the steady state (Eq.~\ref{eq:ss}).  Specifically
In each replicate the initial densities were chosen to be $(XF^*,XH^*,R^*)$, with $X$ a random variable uniformly distributed in $[0,2]$.
By allowing the rate of starvation to vary, we assessed extinction risk across a range of values for $\sigma$ and $\rho$ between ca.\ $10^{-8}$ to $10^{-3}$. %, thus examining a horizontal cross-section of Fig.~\ref{fig:fp}.
Higher rates of extinction correspond to both high values of $\sigma$ if $\rho$ is small, and high values of $\rho$ if $\sigma$ is small.
For low values of $\sigma$ and high values of $\rho$, the increased extinction risk results from higher-amplitude transient cycles as the system nears the Hopf bifurcation (Fig.~\ref{fig:ext}).
For high values of $\sigma$ and low values of $\rho$, increased extinction risk arises because of the decrease in the steady-state consumer population density (Figs. \ref{fig:fp}b, \ref{fig:ext}).
This interplay creates an `extinction refuge', such that for a constrained range of $\sigma$ and $\rho$, extinction probabilities are minimized.

% \noindent \paragraph*{{\bf Extinction refuge.}} 
We find that the allometrically constrained values of $\sigma$ and $\rho$, each representing different trajectories along the ontogenetic curve (Fig. \ref{fig:growth}), fall squarely within the extinction refuge across a range of $M$ (Fig. \ref{fig:ext}a,b, white points). These values are close enough to the Hopf bifurcation to avoid low steady-state densities, yet distant enough to avoid large-amplitude transient cycles.
That allometric values of $\sigma$ and $\rho$ fall within this relatively small window supports the possibility that a selective mechanism has constrained the physiological conditions driving starvation and recovery rates within populations.
Such a mechanism would select for organism physiology that generates appropriate $\sigma$ and $\rho$ values that minimize extinction risk.
This selection could occur via the tuning of body fat percentages, metabolic rates, and/or biomass maintenance efficiencies.
We also find that as body size increases, the amount of low extinction risk parameter space becomes smaller (Fig. \ref{fig:ext}b), suggesting that the population dynamics for larger organisms are more sensitive to variability in physiological rates.
This finding is in accordance with, and may serve as contributing support for, observations of increased extinction risk among larger mammals \cite{Liow:2008jx}.\\
% Moreover, larger body sizes decrease the steady state resource density, such that fluctuations for larger organisms will be more likely to drive resources to extinction.
% OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
% To summarize, our finding that allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

\noindent \paragraph*{{\bf A mechanism for Cope's rule}} Metabolite transport constraints are widely thought to place strict boundaries on biological scaling~\citep{Brown:1993p708,West:1997cg,Brown:2004wq} and thereby lead to specific predictions on the minimum possible body size for organisms~\citep{West:2002ud}.
Above this bound, a number of energetic and evolutionary mechanisms have been explored to assess the costs and benefits associated with larger body masses, particularly for mammals.
One important such example is the \emph{fasting endurance hypothesis}, which contends that larger body size, with consequent lower metabolic rates and increased ability to maintain more endogenous energetic reserves, may buffer organisms against environmental fluctuations in resource availability~\citep{Millar:1990p923}.
Over evolutionary time, terrestrial mammalian lineages show a significant trend towards larger body size known as Cope's rule~\citep{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}, and it is thought that within-lineage drivers generate selection towards an optimal upper bound of roughly $10^7$ (g)~\citep{Alroy:1998p1594}, a value that is likely limited by higher extinction risk for large taxa over longer timescales~\citep{Clauset:2009fh}.
These trends are thought to be driven by a combination of climate change and niche availability~\citep{Saarinen:2014br}; however the underpinning energetic costs and benefits of larger body sizes, and how they influence dynamics over ecological timescales, have not been explored.
% We argue that the NSM provides a suitable framework to explore these issues.

>>>>

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. 
Of particular importance is the interplay between reproduction and the dynamics of starvation and recovery. % in resource-limited environments
To elucidate this interplay, we introduce a nutritional state-structured model that incorporates two classes of consumer: nutritionally replete, reproducing consumers, and undernourished, non-reproducing consumers. % that are susceptible to mortality
% As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. To elucidate the consequences of this tradeoff, w
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics are typically driven to a steady state. %subject to these constraints between body size and a variety of traits
Moreover, these rates fall within a `refuge' in parameter space, where the probability of population extinction is minimized. 
% Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. 
We also show that our model provides a natural framework to predict
maximum mammalian body size by determining the relative stability of an
otherwise homogeneous population to a competing population with altered percent
body fat. This framework provides a principled mechanism for a within-lineage driver of
Cope's rule.
% For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}

@@

The behavioral ecology of all organisms is influenced by their energetic
states, which directly impacts how they invest reserves in uncertain
environments.  Such behaviors are generally manifested as tradeoffs between
investing in somatic maintenance and growth, or allocating energy towards
reproduction~\citep{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.  The timing of
these behaviors responds to selective pressure, as the choice of the
investment impacts future
fitness~\citep{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.  The influence of
resource limitation on an organism's ability to maintain its nutritional
stores may lead to repeated delays or shifts in reproduction over the course
of an organism's life.

The balance between (a) somatic growth and maintenance, and (b) reproduction depends on resource availability~\citep{Morris:1987eo}.
For example, reindeer invest less in calves born after harsh winters (when the mother's energetic state is depleted) than in calves born after moderate winters~\citep{Tveraa:2003fq}.
Many bird species invest differently in broods during periods of resource scarcity~\citep{Daan:1988va,Jacot:2009dv}, sometimes delaying or even foregoing reproduction for a breeding season~\citep{Martin:1987dl,Stearns:1989ip,Barboza:2002in}.
Even freshwater and marine zooplankton have been observed to avoid reproduction under nutritional stress~\citep{Threlkeld:1976ih}, and those that do reproduce have lower survival rates~\citep{Kirk:1997cc}.
Organisms may also separate maintenance and growth from reproduction over space and time: many salmonids, birds, and some mammals return to migratory breeding grounds to reproduce after one or multiple seasons in resource-rich environments where they accumulate reserves~\citep{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.

@@

Including individual energetic dynamics~\citep{Kooi2000} in a
population-level framework~\citep{Kooi2000,Sousa:2010ez} is
challenging~\citep{Diekmann:2010da}.  A common simplifying approach is the
classic Lotka-Volterra (LV) model, which assumes that consumer population
growth rate depends linearly on resource density~\citep{murdoch:2003}. Here,
we introduce an alternative approach---the Nutritional State-structured Model
(NSM)---that accounts for resource limitation via explicit starvation. In
contrast to the LV model, the NSM incorporates two consumer states: hungry
and full, with only the former susceptible to mortality and only the latter
possessing sufficient energetic reserves to reproduce.  Additionally, we
incorporate allometrically derived constraints on the time scales for
reproduction~\citep{Kempes:2012hy}, starvation, and recovery.  Our model
makes several basic predictions: (i) the dynamics are typically driven to a
refuge far from cyclic behavior and extinction risk, (ii) the steady-state
conditions of the NSM accurately predict the measured biomass densities for
mammals described by Damuth's law \citep{Damuth:1987kr,allen2002,enquist1998,pedersen2017shallow},
  %  energetic equivalence principle, 
  (iii) there is an allometrically constrained upper-bound for mammalian body size, and
  (iv) the NSM provides a selective mechanism for the evolution of larger body size, known as Cope's rule ~\citep{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{fig_FixedPoint2.eps}
\caption{\small{ The transcritical (TC; dashed line) and Hopf bifurcation (solid line) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
  bifurcation conditions separate parameter space into unphysical (left of the TC), cyclic,
  and steady state dynamic regimes.  The colors show the steady state densities for the energetically replete consumers $F^*$.  
  % Steady state densities for the energetically deficient consumers $H^*$ (not shown)
  % scale with those for $F^*$.  
}\label{fig:fp}}
\end{figure}

Consumers transition from the full state $F$ to the hungry state $H$ at a
rate $\sigma$---the starvation rate---and also in proportion to the absence
of resources $(1-R)$ (the maximum resource density has been non
dimensionalized to 1; see SI).  Conversely, consumers recover from state $H$
to state $F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$
represents a ratio between maximal resource consumption and the carrying
capacity of the resource. %and $\approx 2$.
The resources that are eaten by hungry consumers (at rate $\rho R + \delta$)
account for their somatic growth ($\rho R$) and maintenance ($\delta$).  Full
consumers eat resources at a constant rate $\beta$ that accounts for maximal
maintenance and somatic growth (see the SI for mechanistic derivations of
these rates from resource energetics).
%This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
The NSM represents an ecologically motivated fundamental extension of the
idealized starving random walk model of foraging, which focuses on resource
depletion, to include reproduction and resource
replenishment~\citep{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is
a more general formulation than previous models that incorporate
starvation~\citep{Persson:1998hz}.

@@

This system of nondimensional equations follows from a set of first-principle
relationships for resource consumption and growth (see the SI for a full derivation and the dimensional form).
Notice that the total consumer density $F+H$ evolves according to $\dot{F}+\dot{H}=\lambda F-\mu H$.
This resembles the equation of motion for the predator density in the LV model~\citep{murray2011mathematical}, except that the resource density does not appear in the growth term.
% As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.
The rate of reproduction is independent of resource density because the full
consumer partitions a constant amount of energy towards reproduction, whereas
a hungry consumer partitions no energy towards reproduction.  Similarly, the
consumer maintenance terms ($\delta H$ and $\beta F$) are also independent of
resource density because they represent a minimal energetic requirement for
consumers in the $H$ and $F$ state, respectively.

% It follows that model predictions are robust only when $R$ is of the order of 1, which holds for all cases that we explore.

\noindent \paragraph*{ {\bf Steady states of the NSM.}} From the single
internal fixed point (Eq.~\eqref{eq:ss}, see Methods), an obvious constraint
on the NSM is that the reproduction rate $\lambda$ must be less than the
starvation rate $\sigma$, so that the consumer and resource densities are
positive.
%In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ represents a transcritical (TC)
bifurcation~\citep{Strogatz:2008wo} that demarcates a physical from an
unphysical (negative steady-state densities) regime.  The biological
implication of the constraint $\lambda<\sigma$ has a simple
interpretation---the rate at which a macroscopic organism loses mass due to
lack of resources is generally much faster than the rate of reproduction.  As
we will discuss below, this inequality is also a natural consequence of
allometric constraints~\citep{Kempes:2012hy} for organisms within empirically
observed body size ranges. %(Fig.~\ref{fig:gvs})

@@

When the starvation rate $\sigma\gg\lambda$, a substantial fraction of the
consumers are driven to the hungry non-reproducing state.  Because
reproduction is inhibited, there is a low steady-state consumer density and a
high steady-state resource density.  However, if $\sigma/\lambda\to 1$ from
above, the population is overloaded with energetically-replete (reproducing)
individuals, thereby promoting transient oscillations between the consumer
and resource densities (Fig.~\ref{fig:fp}).  If the starvation rate is low
enough that the Hopf bifurcation is crossed, these oscillations become
stable.  This threshold occurs at higher values of the starvation rate as the recovery rate $\rho$ increases, such that the range of parameter space giving rise to cyclic dynamics also increases with higher recovery rates.\\

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{Growth-trajectory-diagram.pdf}
\caption{\small{ The growth trajectory over absolute time of an individual organism as a function of body mass.  
Initial growth follows the black trajectory to an energetically replete reproductive adult mass of $m=\epsilon_\lambda M$ (see Methods). %which we assume is 95\% asymptotic mass $M$.  
Starvation follows the red trajectory to $m = \epsilon_\sigma \epsilon_\lambda  M$. 
Recovery follows the green curve to the replete adult mass, where this trajectory differs from the original growth because only fat is being regrown which requires a longer time to reach $\epsilon_\lambda M$. %different energetics and 
Alternatively, death from starvation follows the blue trajectory to $m=\epsilon_\mu \epsilon_\lambda  M$.}\label{fig:growth}}
\end{figure}

% \noindent {\bf Role of allometry} \\
%[Link Allometry stuff to our model - what does it provide to the story]
%[Introduce biological and linked constraints]
%[Allometries capture vast amounts of diversity via a single parameter --- body size]
%[Allometries have captured intterest etc across multiple scales and biological classes of organisms]
\noindent {\bf Results}
\noindent \paragraph*{{\bf The allometry of extinction risk.}} While there
are no {\it a priori} constraints on the parameters in the NSM, we expect
that each species should be restricted to a distinct portion of the parameter
space.  We use allometric scaling relations to constrain the covariation of
rates in a principled and biologically meaningful manner (see Methods).
Allometric scaling relations highlight common constraints and average trends
across large ranges in body size and species diversity. Many of these
relations can be derived from a small set of assumptions.  In the Methods we
describe our framework to determine the covariation of timescales and rates
across a range of body sizes for each of the key parameters of our model
(cf.\ Ref.~\citep{Yodzis:1992hg}).
% We are thereby able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

Nearly all of the rates described in the NSM are determined by consumer
metabolism, which can be used to describe a variety of organismal features
\citep{Brown:2004wq}.  We derive, from first principles, the relationships
for the rates of reproduction, starvation, recovery, and mortality as a
function of an organism's body size and metabolic rate (see Methods).
Because we aim to explore the starvation-recovery dynamics as a function of
an organism's body mass $M$, we parameterize these rates in terms of the
\emph{percent} gain and loss of the asymptotic (maximum) body mass,
$\epsilon M$, where different values of $\epsilon$ define different states of
the consumer (Fig.~\ref{fig:growth}; see Methods for derivations of
allometrically constrained rate equations).  Although the rate equations
\eqref{eq:system} are general and can in principle be used to explore the
starvation recovery dynamics for most organisms, here we focus on allometric
relationships for terrestrial-bound lower-trophic level endotherms (see the
SI for values), specifically herbivorous mammals, which range from a minimum
of $M\approx1$g (the Etruscan shrew \emph{Suncus etruscus}) to a maximum of
$M\approx10^7$g (the early Oligocene Indricotheriinae and the Miocene
Deinotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associated with
$\epsilon$. Moreover, we emphasize that our allometric equations (see
Methods) describe mean relationships, and do not account for the (sometimes
considerable) variance associated with individual species.  We note that
including additional allometrically-scaled mortality terms to both $F$ and
$H$ does not change the form of our model nor impact our quantitative findings
(see SI for the derivation).

@@

%Extinction to the left and the right
% \noindent \paragraph*{{\bf Extinction risk}} 
As the allometric derivations of the NSM rate laws reveal (see Methods),
starvation and recovery rates are not independent parameters, and the
biologically relevant portion of the phase space shown in Fig.~\ref{fig:fp}
is constrained via covarying parameters.  Given the parameters of terrestrial
endotherms, we find that the starvation rate $\sigma$ and the recovery rate
$\rho$ are constrained to lie within a small region of potential values for
the known range of body sizes $M$.  Indeed, starvation and recovery rates
across all values of $M$ fall squarely in the steady-state region at some
distance from the Hopf bifurcation.  This suggests that cyclic population
dynamics should be rare, particularly in resource-limited environments.
% Moreover, the dynamics for all mammalian body sizes are confined to the steady-state regime of the NSM and that limit-cycle behavior is precluded.

Higher rates of starvation result in a larger flux of the population to the hungry state.
In this state, reproduction is absent, thus increasing the likelihood of extinction.  From the perspective of population survival, it is the rate of starvation relative to the rate of recovery that determines the long-term dynamics of the various species (Fig.~\ref{fig:fp}).
We therefore examine the competing effects of cyclic dynamics vs.\ changes in steady-state density on extinction risk, both as functions of $\sigma$ and $\rho$.
To this end, we computed the probability of extinction, where we define extinction as a population trajectory falling below one fifth of the allometrically constrained steady state at any time between $t=10^8$ and $t=10^{10}$.
This procedure was repeated for 50 replicates of the continuous-time system shown in Eq.~\ref{eq:system} for organisms with mass ranging from $10^2$ to $10^6$ grams.
%the initial condition isdistributed around the steady state (Eq.~\ref{eq:ss}).  Specifically
In each replicate the initial densities were chosen to be $(XF^*,XH^*,R^*)$,
with $X$ a random variable uniformly distributed in $[0,2]$.  By allowing the
rate of starvation to vary, we assessed extinction risk across a range of
values for $\sigma$ and $\rho$ between ca.\ $10^{-8}$ to
$10^{-3}$. %, thus examining a horizontal cross-section of Fig.~\ref{fig:fp}.
Higher rates of extinction correspond to both large $\sigma$ if $\rho$ is
small, and large $\rho$ if $\sigma$ is small.  In the former case, increased
extinction risk arises because of the decrease in the steady-state consumer
population density (Figs. \ref{fig:fp}b, \ref{fig:ext}).  In the latter case,
the increased extinction risk results from higher-amplitude transient cycles
as the system nears the Hopf bifurcation (Fig.~\ref{fig:ext}).  This
interplay creates an `extinction refuge', such that for a constrained range
of $\sigma$ and $\rho$, extinction probabilities are minimized.

% \noindent \paragraph*{{\bf Extinction refuge.}} 
We find that the allometrically constrained values of $\sigma$ and $\rho$,
each representing different trajectories along the ontogenetic curve
(Fig. \ref{fig:growth}), fall squarely within the extinction refuge across a
range of $M$ (Fig. \ref{fig:ext}a,b, white points). These values are close
enough to the Hopf bifurcation to avoid low steady-state densities, yet
distant enough to avoid large-amplitude transient cycles.  Allometric values
of $\sigma$ and $\rho$ fall within this relatively small window, which
supports the possibility that a selective mechanism has constrained the
physiological conditions driving starvation and recovery rates within
populations.  Such a mechanism would select for organism physiology that
generates appropriate $\sigma$ and $\rho$ values that minimize extinction
risk.  This selection could occur via the tuning of body fat percentages,
metabolic rates, and/or biomass maintenance efficiencies.  We also find that
as body size increases, the size of the low extinction-risk parameter space
shrinks (Fig.~\ref{fig:ext}b), suggesting that the population dynamics for
larger organisms are more sensitive to variability in physiological rates.
This finding is in accordance with, and may serve as contributing support for, observations of increased extinction risk among larger mammals \cite{Liow:2008jx}.\\
% Moreover, larger body sizes decrease the steady state resource density, such that fluctuations for larger organisms will be more likely to drive resources to extinction.
% OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
% To summarize, our finding that allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

\noindent \paragraph*{{\bf A mechanism for Cope's rule}} Metabolite transport
constraints are widely thought to place strict boundaries on biological
scaling~\citep{Brown:1993p708,West:1997cg,Brown:2004wq} and thereby lead to
specific predictions on the minimum possible body size for
organisms~\citep{West:2002ud}.  Above this bound, a number of energetic and
evolutionary mechanisms have been explored to assess the costs and benefits
associated with larger body masses, particularly for mammals.  One important
such example is the \emph{fasting endurance hypothesis}, which contends that
larger body size, with consequent lower metabolic rates and increased ability
to maintain more endogenous energetic reserves, may buffer organisms against
environmental fluctuations in resource availability~\citep{Millar:1990p923}.
Over evolutionary time, terrestrial mammalian lineages show a significant
trend towards larger body size---Cope's
rule~\citep{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}.
It is thought that within-lineage drivers generate selection towards an
optimal upper bound of roughly $10^7$ (g)~\citep{Alroy:1998p1594}, a value
that is likely limited by higher extinction risk for large taxa over longer
timescales~\citep{Clauset:2009fh}.  These trends are thought to be driven by
a combination of climate change and niche
availability~\citep{Saarinen:2014br}; however the underpinning energetic
costs and benefits of larger body sizes, and how they influence dynamics over
ecological timescales, have not been explored.
% We argue that the NSM provides a suitable framework to explore these issues.
